Titeln på avhandlingens språk

### 2023-05-19 body/abstract.tex
cleaning up for publication

\phantomsection\addcontentsline{toc}{chapter}{Abstract}
\begin{abstract}
    Cloud Computing, with its globally-accessible nature and virtually unlimited scalability, has revolutionized our daily lives since its widespread adoption in the early 2000s.
    It allows us to access our documents anywhere, keep in contact with our friends, back up our photos, and even remote-control some of our appliances.
    Despite this, Cloud Computing has limitations when it comes to novel applications requiring real-time processing or low-latencies.
    Applications such as \glspl{CPS} or mobile \gls{XR}, which in turn also have great transformative potential, are unable to run on the Cloud.

    Edge Computing is emerging as a potential solution to these limitations by bringing computation and data processing closer to the edge of the network, thereby reducing latency and enabling real-time decision making.
    The combination of Edge Computing and modern mobile network technologies such as 5G offers potential for massive deployments of latency-sensitive applications.
    However, scaling and understanding these deployments poses important challenges such the optimization of latency through multiple processing steps and trade-offs in wireless system choice, protocols, hardware, and algorithms.
    Existing approaches have so far been unsuccessful in capturing the complex effects arising from the interplay between network and compute in these systems.

    This dissertation addresses the challenge of performance evaluation of Edge Computing and the applications enabled by this paradigm with two key contributions to literature.
    First, a methodological approach to experimentally studying the trade-offs between system responsiveness and resource consumption in Edge-bound latency-sensitive applications such as \glspl{CPS} and \gls{XR} is introduced.
    These applications and systems feature characteristics, such as tight interaction with the physical world and the involvement of humans, that make them challenging to study through simulated approaches or analytical modeling.
    The approach presented in this thesis involves therefore the emulation of the client-side workload while maintaining the real server-side process and network stack to retain the realism of network and compute effects.

    Next, an exploration of the requirements for accuracy in the emulation is presented.
    Further, this work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
    To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
    The model is combined with a mathematical framework to study the potential for optimization in these applications.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% SWEDISH ABSTRACT %%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\selectlanguage{swedish}
\phantomsection\addcontentsline{toc}{section}{Abstract (Swedish)}
\glsresetall%
\begin{abstract}
    Cloud Computing, med sin globalt tillgängliga natur och praktiskt taget obegränsade skalbarhet, har revolutionerat våra dagliga liv sedan dess allmänna antagande i början av 2000-talet.
    Det gör det möjligt för oss att få tillgång till våra dokument var som helst, hålla kontakt med våra vänner, säkerhetskopiera foton, och till och med fjärrstyra våra hushållsapparater. 
    Trots detta har Cloud Computing begränsningar när det gäller nya applikationer som kräver realtidsbehandling eller låg latens.
    Applikationer som cyberfysiska system (engelska \emph{\gls{CPS}}) eller mobilt utvidgad verklighet (engelska \emph{\gls{XR}}), som i sin tur också har stor omvandlingspotential, kan inte köras på molnet.

    Edge Computing är en potentiell lösning på dessa begränsningar som, genom att föra beräkningar och databehandling närmare kanten av nätverket, minskar latensen och möjliggör beslutsfattande i realtid.
    Kombinationen av Edge Computing och moderna mobila nätverksteknologier som 5G erbjuder potential för massiva utrullningar av latenskänsliga applikationer. 
    Men att skala och förstå dessa utrullningar utgör viktiga utmaningar såsom optimering av latens genom flera bearbetningssteg och avvägningar i val av trådlösa system, protokoll, hårdvara och algoritmer.
    Befintliga metoder har hittills inte lyckats fånga de komplexa effekter som uppstår från samspelet mellan nätverk och datorer i dessa system.

    Denna avhandling tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
    Först introduceras ett metodologiskt tillvägagångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
    Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillvägagångssätt eller analytisk modellering.
    Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att bibehålla realismen i nätverks- och datoreffekter.

    Därefter presenteras en utforskning av kraven på noggrannhet i emuleringen.
    Vidare diskuterar detta arbete i vilken utsträckning noggrannhet i emuleringen kan öppna nya vägar för optimering av dessa system.
    För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobilt förstärkt verklighets-applikationer (engelska \emph{\gls{MAR}}).
    Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i dessa applikationer.

    Resultaten indikerar att den metod som introducerats i detta arbete erbjuder fördelar jämfört med befintliga metoder genom att förbättra effektiviteten, repeterbarheten och replikerbarheten.
    Genom att helt integrera arbetsbelastningskomponenter i den emulerade mjukvarudomänen, minskar denna metodik komplexiteten samtidigt som den fångar komplexa effekter av nätverks- och beräkningsfaktorer som är utmanande att modellera.
    Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
    Dessutom tyder resultat från utforskningen av implikationerna av noggrannhet i emuleringen att inkorporering av förbättrad noggrannhet i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
    Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

>>>>

\phantomsection\addcontentsline{toc}{chapter}{Abstract}
\begin{abstract}
    Cloud Computing, with its globally-accessible nature and virtually unlimited scalability, has revolutionized our daily lives since its widespread adoption in the early 2000s.
    It allows us to access our documents anywhere, keep in touch with friends, back up photos, and even remotely control our appliances.
    Despite this, Cloud Computing has limitations when it comes to novel applications requiring real-time processing or low-latencies.
    Applications such as \glspl{CPS} or mobile \gls{XR}, which in turn also have great transformative potential, are unable to run on the Cloud.

    Edge Computing is emerging as a potential solution to these limitations by bringing computation closer to the edge of the network, thereby reducing latency and enabling real-time decision making.
    The combination of Edge Computing and modern mobile network technologies such as 5G offers potential for massive deployments of latency-sensitive applications.
    However, scaling and understanding these deployments poses important challenges such the optimization of latency through multiple processing steps and trade-offs in wireless system choice, protocols, hardware, and algorithms.
    Existing approaches have so far been unsuccessful in capturing the complex effects arising from the interplay between network and compute in these systems.

    This dissertation addresses the challenge of performance evaluation of Edge Computing and the applications enabled by this paradigm with two key contributions to literature.
    First, a methodological approach to experimentally studying the trade-offs between system responsiveness and resource consumption in latency-sensitive applications such as \glspl{CPS} and \gls{XR} is introduced.
    These applications and systems feature characteristics, such as tight interaction with the physical world and the involvement of humans, that make them challenging to study through simulated approaches or analytical modeling.
    The approach presented in this thesis involves therefore the emulation of the client-side workload while maintaining the real server-side process and network stack to retain the realism of network and compute effects.

    Next, an exploration of the requirements for accuracy in the emulation is presented.
    This work discusses the extent to which accuracy in the emulation can open new avenues for optimization of these systems.
    To this end, the first-ever realistic model of human timings for a particular class of \gls{MAR} applications is provided. 
    The model is combined with a mathematical framework to study the potential for optimization in Edge Computing applications.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% SWEDISH ABSTRACT %%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\selectlanguage{swedish}
\phantomsection\addcontentsline{toc}{section}{Abstract (Swedish)}
\glsresetall%
\begin{abstract}
    Cloud Computing, med sin globalt tillgängliga natur och praktiskt taget obegränsade skalbarhet, har revolutionerat våra dagliga liv sedan dess allmänna antagande i början av 2000-talet.
    Det gör det möjligt för oss att få tillgång till våra dokument var som helst, hålla kontakt med vänner, säkerhets-kopiera foton, och fjärrstyra våra hushållsapparater. 
    Trots detta har Cloud Computing begränsningar när det gäller nya applikationer som kräver realtidsbehandling eller låg latens.
    Applikationer som cyberfysiska system (engelska \emph{\gls{CPS}}) eller mobilt utvidgad verklighet (engelska \emph{\gls{XR}}), som i sin tur också har stor omvandlingspotential, kan inte köras på molnet.

    Edge Computing framstår som en potentiell lösning på dessa begräns-ningar genom att föra beräkningar närmare kanten av nätverket, vilket minskar latensen och möjliggör beslutsfattande i realtid.
    Kombinationen av Edge Computing och moderna mobila nätverksteknologier som 5G erbjuder potential för massiva utrullningar av latenskänsliga applikationer. 
    Men att skala och förstå dessa utrullningar utgör viktiga utmaningar såsom optimering av latens genom flera bearbetningssteg och avvägningar i val av trådlösa system, protokoll, hårdvara och algoritmer.
    Befintliga metoder har hittills inte lyckats fånga de komplexa effekter som uppstår från samspelet mellan nätverk och datorer i dessa system.

    Denna avhandling tar upp utmaningen med prestationsutvärdering av Edge Computing och applikationerna som möjliggörs av detta paradigm med två viktiga bidrag till litteraturen.
    Först introduceras ett metodologiskt tillväga-gångssätt för att experimentellt studera kompromisserna mellan systemrespons och resursförbrukning i latenskänsliga applikationer såsom \glspl{CPS} och \gls{XR} som körs på Edge Computing.
    Dessa applikationer och system har egenskaper, såsom tät interaktion med den fysiska världen och inblandning av människor, som gör dem utmanande att studera genom simulerade tillväga-gångssätt eller analytisk modellering.
    Den metod som presenteras i denna avhandling innebär därför emulering av klientsidans arbetsbelastning samtidigt som den verkliga serversidans process och nätverksstacken bibehålls för att behålla realismen i nätverks- och datoreffekter.

    Därefter presenteras en utforskning av kraven på precision i emuleringen.
    Detta arbete diskuterar i vilken utsträckning precision i emuleringen kan öppna nya vägar för optimering av dessa system.
    För detta ändamål tillhandahålls den första realistiska modellen av mänskliga tidsbeteenden för en särkild klass av mobil förstärkt verklighet (engelska \emph{\gls{MAR}}).
    Modellen kombineras med ett matematiskt ramverk för att studera potentialen för optimering i Edge Computing applikationer.

    Resultaten indikerar att den metod som introducerats i detta arbete erbjuder fördelar jämfört med befintliga metoder genom att förbättra effektiviteten, repeterbarheten och replikerbarheten.
    Genom att helt integrera arbetsbelastningskomponenter i den emulerade mjukvarudomänen, minskar denna metodik komplexiteten samtidigt som den fångar komplexa effekter av nätverks- och beräkningsfaktorer som är utmanande att modellera.
    Detta tillvägagångssätt representerar således ett viktigt bidrag till litteraturen, eftersom det består av en omfattande metod för prestandautvärdering av Edge Computing-miljöer, som omfattar både applikationen och infrastrukturen.
    Dessutom tyder resultat från utforskningen av implikationerna av precision i emuleringen att inkorporering av förbättrad precision i klientsideemulering kan möjliggöra implementering av optimeringsmetoder som annars skulle vara omöjliga.
    Särskilt framhåller detta arbete betydelsen av att överväga mänskligt beteende och reaktioner utöver systemrelaterade mätvärden och prestandaoptimeringar i samband med \gls{MAR}.

### 2023-05-19 body/acks.tex
cleaning up for publication

I would also like to thank everybody involved with the defense of this thesis.
I am very grateful to Prof. Yu Xiao of Aalto University, for agreeing to act as my dissertation opponent.
My dissertation defense will certainly benefit greatly from her expertise and insights.
I would like to extend my sincere thanks to my grading committe, consisting of Prof. Ana Aguiar (Universidade do Porto), Prof. Per Gunningberg (Uppsala Universitet), and Prof. Klaus Wehrle (\gls{RWTH}), for investing time and effort into reading and grading this dissertation.
Special thanks also go to Prof. Markus Flierl of \pgls{KTH} for advance-reviewing this dissertation, and to Prof. Mats Bengtsson, also of \pgls{KTH}, for agreeing to act as chairperson for the defense.

@@

I am very lucky and grateful that I got to stay at both \pgls{KTH} and \gls{CMU}, and I would like to thank everyone who made these spaces welcoming and engaging.
A big thank you goes out to my colleagues (both current and former) at the division of \gls{ISE} for all those fun lunches at Restaurang Q, Cypern, and SysterOBror over the years, which always helped to break the monotony of the days.
This includes but is not limited to Samie, Vishnu, Neel, Sahar, Antonios, Germán, Boules, Baptiste, Håkan, Pol, Marie, Wendi, and Sebastian.
I would additionally like to extend a special thank you to Dr. Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first moths.
Likewise, I extend my gratitute to Gerd Fransson, for greeting me with a huge smile every morning the years our offices faced each other and cheering up my days with her wholesome energy.
I also thank Prof. Joakim Jaldén for giving me the opportunity to act as his teaching assistant.
Another big thank you goes out to everyone (current and former) at Satya's group at \gls{CMU}, for always welcoming me with open arms and treating me like one of them whenever I was visiting.
This includes but is not limited to Junjue, Tom, Roger, Shilpa, Jan, Edmond, Jim, Babu, and Zhuo.
I will always remember our Monday morning talks and Friday lunch outings dearly.

@@

This dissertation would not have been possible without the help and support of my family.
First and foremost, I would like to thank the love of my life and wife, Andresa.
I could not have done this without her companionship and love.
I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
They have been my rocks, my pillars, since the day I was born, and I am proud to be their son.
A big thank you goes out to my sister, Paola, as well.
Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
Likewise, I extend my deepest gratitude to everyone in my immediate and extended family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.
I hope he would have been proud seeing how far I have come.

>>>>

I would also like to thank everybody involved with the defense of this thesis.
I am very grateful to Prof. Yu Xiao of Aalto University, for agreeing to act as my dissertation opponent.
My defense will certainly benefit greatly from her expertise and insights.
I would like to extend my sincere thanks to my grading committe, consisting of Prof. Ana Aguiar (Universidade do Porto), Prof. Per Gunningberg (Uppsala Universitet), and Prof. Klaus Wehrle (\gls{RWTH}), for investing time and effort into reading and grading this dissertation.
Special thanks also go to Prof. Markus Flierl of \pgls{KTH} for advance-reviewing this work, and to Prof. Mats Bengtsson, also of \pgls{KTH}, for agreeing to act as chairperson for the defense.

@@

I am very lucky and grateful that I got to stay at both \pgls{KTH} and \gls{CMU}, and I would like to thank everyone who made these spaces welcoming and engaging.
A big thank you goes out to my colleagues (both current and former) at the division of \gls{ISE} for all those fun lunches at Restaurang Q, Cypern, and SysterOBror over the years, which always helped to break the monotony of the days.
This includes but is not limited to Samie, Vishnu, Neel, Sahar, Antonios, Germán, Boules, Baptiste, Håkan, Pol, Marie, Wendi, and Sebastian.
I would additionally like to extend a special thank you to Dr. Sebastian Schiessl, for ``taking me under his wing'' when I first arrived at \gls{ISE} and acting as a sort of mentor during my first months.
Likewise, I extend my gratitute to Gerd Fransson, for greeting me with a huge smile every morning the years our offices faced each other and cheering up my days with her wholesome energy.
I also thank Prof. Joakim Jaldén for giving me the opportunity to act as his teaching assistant.
Another big thank you goes out to everyone (current and former) at Satya's group at \gls{CMU}, for always welcoming me with open arms and treating me like one of them whenever I was visiting.
This includes but is not limited to Junjue, Tom, Roger, Shilpa, Jan, Edmond, Jim, Babu, and Zhuo.
I will always remember our Monday morning talks and Friday lunch outings dearly.

@@

This dissertation would not have been possible without the help and support of my family.
First and foremost, I would like to thank the love of my life and wife, Andresa.
I could not have done this without her companionship and love.
I am eternally grateful for her unconditional support over difficult periods of time and over tremendous distances while we were living apart, and for making every day we spend together amazing.
I am also deeply grateful to my parents, Gabriel and Valeria, for always encouraging my curiosity and my drive to learn and discover.
They have been my rocks, my pillars, since the day I was born, and I am proud to be their son.
A big thank you goes out to my sister, Paola, as well.
Although we sometimes get on each other's nerves, she has always been there for me and supported me, and I could not have asked for a better sister.
Likewise, I extend my deepest gratitude to everyone in my family who have given me support and love over the years, including but not limited to my paternal grandparents, Manuel and María Yolanda, my maternal grandparents, Caupolicán and María Aura, my uncle Mauricio, his wife Marylen, their daughter Amanda, and everybody else.
I would like to make a special mention to my late grandfather, Caupolicán ``Poli'' Muñoz, who always inspired and fanned the flames of academic curiosity in me.
I hope he would have been proud seeing how far I have come.

### 2023-05-19 body/part1/chapter1-introduction/background.tex
cleaning up for publication

In this section, we will discuss two key examples of such systems, \glspl{CPS} and \gls{XR} applications.
\glspl{CPS} are systems that integrate physical and computational components, while \gls{XR} applications involve the creation of immersive virtual environment that combine the physical and virtual worlds.
Both types of systems require real-time monitoring and control of complex processes through feedback loops of either sensor inputs and control outputs in the case of \glspl{CPS}, and video and image streams in the case of \gls{XR}.
This section will introduce the key characteristics of these systems, providing a thorough background to understand the opportunities and challenges they present.

@@

In this dissertation, we focus specifically on \emph{step-based} \gls{WCA}.
These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users, illustrated in \cref{fig:wca:operation}.
The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
The application follows the progress of the task in real-time by repeatedly sampling the state of the physical system, most commonly through video feeds.
Whenever the assistant detects a change in the context of the user, it provides a new instruction.
The application otherwise remains silent and out-of-the-way; samples with unfinished states or noise are silently discarded.

>>>>

In this section, I will discuss two key examples of such systems, \glspl{CPS} and \gls{XR} applications.
\glspl{CPS} are systems that integrate physical and computational components, while \gls{XR} applications involve the creation of immersive virtual environment that combine the physical and virtual worlds.
Both types of systems require real-time monitoring and control of complex processes through feedback loops of either sensor inputs and control outputs in the case of \glspl{CPS}, and video and image streams in the case of \gls{XR}.
This section will introduce the key characteristics of these systems, providing a thorough background to understand the opportunities and challenges they present.

@@

In this dissertation, I focus specifically on \emph{step-based} \gls{WCA}.
These applications operate in a manner analogous to how \gls{GPS} navigation assistants guide users, illustrated in \cref{fig:wca:operation}.
The progress of the user is seamlessly and continuously monitored by the application, which provides relevant instructions and feedback only when these are needed.
The application follows the progress of the task in real-time by repeatedly sampling the state of the physical system, most commonly through video feeds.
Whenever the assistant detects a change in the context of the user, it provides a new instruction.
The application otherwise remains silent and out-of-the-way; samples with unfinished states or noise are silently discarded.

### 2023-05-19 body/part1/chapter1-introduction/introduction.tex
cleaning up for publication

However, this design leads to trade-offs in bandwidth and, particularly, latency.
Cloud datacenter are commonly located in geographically centralized locations to better serve thousands or even millions of users.
Conversely, this results in these datacenters having a large average \emph{topological} and \emph{physical} distance to individual users.
Topologically, end-user connections to the Cloud tend to transit a significant number of hops, i.e.\ intermediate connections through networking equipment belonging to service providers at multiple tiers of the internet.
Data needs to be received, routed, and re-transmitted at each hop, adding delay.
Physically, Cloud datacenters are often located hundreds or even thousands of kilometers from end-users.
Connections are limited in their responsiveness by the speed of light, and these long physical distances directly translate into higher latencies.

@@

\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive application systems.
\gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
\gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
An  example of such an application is illustrated in \cref{fig:augmentedreality}.
By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.

@@

The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
However, these systems exhibit several characteristic that make their study complex.
One of these is unique nature of Edge computing environments, which tightly integrates compute and network.
This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
Such expertise may not always be feasible to obtain within a single research group.

@@

The scalability of these systems is also a challenge.
It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, personnel and test subjects, as well as facilities.
\glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{XR} is time-consuming and expensive.
This also makes the study of these applications hard to automate, making it challenging to conduct large-scale experiments or to collect data over an extended period of time.

This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects (e.g.~\cite{chen2015efficient} and~\cite{al2017reliable}), resulting in potential trade-offs that may neglect one half of the larger picture.
We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.

This dissertation aims to contribute to the field by presenting two key contributions.
First, we introduce and subsequently investigate the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
The proposed methodology is based on the emulation of workloads and aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
Next, we present a deeper exploration of the implications of improved realism in the emulated components of our methodology by studying potential avenues for optimization of a particular sub-class of \gls{MAR}.

Our methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{XR} applications. 
We emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
    In our initial implementation, these correspond to low-cost and easily replaceable, scalable Raspberry Pi 4 Model B \acsp{SBC}.
} running a realistic software imitation of the desired behaviors.
At the same time, our methodology retains the real network as well as the real compute hardware and software at the backend.
This design results in real-time benchmarking and performance evaluation of applications and Edge infrastructure, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.

Emulating the workload component reduces complexity by moving it into the software domain, allowing for improved repeatability and replicability, as well as easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
This also reduces the barrier of entry to this research, as \glspl{SBC} are for the most part cheap and easily accessible.
Additionally, it preserves the realism of effects stemming from the hardware and network.
In particular, the methodology allows us to capture effects due to the interaction of network factors such as contention, congestion control, and medium access, with effects arising due to process scheduling and/or resource contention at the compute level.
These effects are often of stochastic or chaotic natures and complex to capture in models or simulations.

Next, we study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
To achieve this, we develop a novel model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
We then combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications.
We find that improved accuracy in the emulation enables otherwise inviable optimization approaches.

We believe our contributions represent valuable tools and insights for researchers, system designers, and application developers.
The methodology represents a novel, holistic approach to the performance evaluation of Edge environments.
Furthermore, our findings have significant implications for the design and optimization of applications enabled by Edge Computing, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
We hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing and 5G systems.

>>>>

However, this design leads to trade-offs in bandwidth and, particularly, latency.
Cloud datacenter are commonly located in geographically centralized locations to better serve thousands or even millions of users.
Conversely, this results in these datacenters having a large average \emph{topological} and \emph{physical} distance to individual users.
Topologically, end-user connections to the Cloud tend to transit a significant number of hops, i.e.\ intermediate connections through networking equipment belonging to service providers at multiple tiers of the internet.
Data needs to be received, routed, and re-transmitted at each hop, adding delay.
Physically, Cloud datacenters are often located hundreds or even thousands of kilometers from end-users.
Connections are limited in their responsiveness by the speed of light, and long physical distances directly translate into higher latencies.

@@

\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive applications.
\gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}).
\gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
An  example of such an application is illustrated in \cref{fig:augmentedreality}.
By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.

@@

The evaluation of performance in Edge Computing has been addressed in existing works through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
However, these systems exhibit characteristic that make their study complex.
One of these is the unique nature of Edge computing environments, which tightly integrates compute and network.
This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, and queuing analysis.
Such expertise may not always be feasible to obtain within a single research group.

@@

The scalability of these systems is also a challenge.
It may not be feasible for researchers to test the applications on a large scale due to limitations in access to specialized equipment, test subjects, as well as facilities.
\glspl{CPS} often rely on physical sensors and actuators, making it difficult to replicate real-world conditions in a controlled environment at scale, and conducting experiments involving human subjects for \gls{XR} is time-consuming and expensive.
This also makes the study of these applications hard to automate, making it challenging to conduct large-scale experiments or to collect data over an extended period of time.

This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects (e.g.~\cite{chen2015efficient} and~\cite{al2017reliable}), resulting in potential trade-offs that may neglect one half of the larger picture.
I argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.

This dissertation aims to contribute to the field by presenting two key contributions.
First, I introduce and subsequently investigate the applications of a methodology for the study of latency-sensitive applications deployed on Edge environments.
The proposed methodology is based on the emulation of workloads and aims to enhance the accuracy and realism of results related to Edge infrastructure and 5G networks, particularly in regard to network performance.
Next, I present a deeper exploration of the implications of improved accuracy in the emulated components of my methodology by studying potential avenues for optimization of a particular sub-class of \gls{MAR}.

My methodology is designed to tackle these challenges by employing an emulation approach to the benchmarking of latency-sensitive \gls{CPS} and \gls{XR} applications. 
I emulate target workloads on actual Edge infrastructure by replacing the client side of the system with \gls{COTS} general-purpose computing devices\footnote{%
    In my initial implementation, these correspond to low-cost and easily replaceable, scalable Raspberry Pi 4 Model B \acsp{SBC}.
} running a realistic software imitation of the desired behaviors.
At the same time, my methodology retains the real network as well as the real compute hardware and software at the backend.
This design results in real-time benchmarking and performance evaluation of applications and Edge infrastructure, while still providing realism, scalability, and reduced complexity, as well as allowing for easier automation of these studies.

Emulating the workload component reduces complexity by moving it into the software domain, allowing for improved repeatability and replicability, as well as easier horizontal scaling through the use of \gls{COTS} general-purpose hardware such as \glspl{SBC}.
This also reduces the barrier of entry to this research, as \glspl{SBC} are for the most part cheap and easily accessible.
Additionally, it preserves the realism of effects stemming from the hardware and network.
In particular, the methodology allows for the capture of effects due to the interaction of network factors such as contention, congestion control, and medium access, with factors due to process scheduling and/or resource contention at the compute level.
These effects are often of stochastic or chaotic natures and complex to capture in models or simulations.

Next, I study how improved accuracy in the emulations employed by this methodology can open new avenues for the optimization of applications on Edge Computing.
To achieve this, I develop a model for human behavior in a subclass of \gls{MAR} which is able to dynamically adapt to changes in system responsiveness.
I combine it with a mathematical framework to study the potential for optimization of resource consumption in these applications, and find that improved accuracy in the emulation enables otherwise inviable optimization approaches.

I believe my contributions represent valuable tools and insights for researchers, system designers, and application developers.
The methodology represents a novel, holistic approach to the performance evaluation of Edge environments.
Furthermore, my findings have significant implications for the design and optimization of applications enabled by Edge Computing, particularly in the areas of energy consumption, sampling strategies, and application lifetime.
I hope and believe this dissertation will contribute to the development of new techniques and approaches for improving the performance and reliability of Edge Computing.

### 2023-05-19 body/part1/chapter1-introduction/structure.tex
cleaning up for publication

This dissertation is structured into two parts.
\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
Next, in \cref{part:publications} we present the publications that form the core of this thesis.

\cref{part:summary} is itself structured into four main chapters.
\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
We aim to demonstrate the significance of our work and how it contributes to the field.
Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

>>>>

This dissertation is structured into two parts.
\cref{part:summary} presents a summary, introducing the topic, discussing related work, and highlighting key contributions.
Next, in \cref{part:publications} I present the publications that form the core of this thesis.

\cref{part:summary} is itself structured into four main chapters.
\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
In \cref{chap:relwork}, I present the scope of the thesis, as well as relevant related work.
The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
I discuss context for my research, and establish the foundation for the contributions presented in the dissertation.
\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
I discuss methods, tools, and findings that were developed and obtained throughout the course of my research.
I aim to demonstrate the significance of my work and how it contributes to the field.
Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
I summarize the key findings and contributions of the thesis, and briefly discuss the implications of my findings and contributions for the field.
This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.
